Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-12-28 manuscript/manuscript.tex
Edits for the linear density case
Moved some things down to the discussion and deleted the parts that I
moved up to the previous section.

\noindent
has an analytical solution given by Eq.~\ref{eq:shell-pot-linear}.

The absolute density difference defined on equation~\ref{eq:density-abs-diff} will
always be zero for the linear density case.
As a result, the inequality~\ref{eq:delta-density} will always be satisfied and no
divisions will ever be performed.
Therefore, the two dimensional adaptive discretization algorithm is the
only mechanism that controls the accuracy of the numerical integration.
For this reason, we will only determine the minimum value of $D$ needed in order to
guarantee an acceptable accuracy while ignoring the value of $\delta$.

In order to compare the numerical results with the analytical solution we must build
spherical shell models made of tesseroids: we choose to regularly split the spherical
shell along the latitudinal and longitudinal directions obtaining a shell model made out
of $30^\circ \times 30^\circ$ tesseroids.
In order to cover different computation scenarios we use five spherical shells, each one
with the outer radius equal to the mean Earth radius and the inner radius determined by
the desired thickness (100m, 1km, 10km, 100km and 1000km).
These shell models are detailed on Table~\ref{tab:shell-models}.

The comparison between the numerical results and the analytical solution could be
naively compared on a single computation point due to the rotation symmetry of the
spherical shell.
But the numerical results may adopt different values depending on the relative location
between the computation point and the tesseroid, thus is not wise to carry out the
comparison on a single point.
Instead, we propose four computation grids detailed on Table~\ref{tab:grids}: a Pole and
an Equator grids, and a Global and Satellite grids.
The former ones are regional scale grids located on the outer surface of the spherical
shell, while the latter are global scale grids located at zero height (Global) and at
the mean orbital height of the gravimetric satellite missions (Satellite).
These grids are defined in the spirit of spanning a broad scenario of gravitational
field computations and ensure an acceptable accuracy on each one of them.
The comparison will be carried out by computing the maximum absolute difference between
the analytical solution and the numerical result on each point of the computation grid.

We assign a density that varies linearly with $r$ (Eq.~\ref{eq:density-linear}) to each
shell model, which assumes the values of $\rho_\text{in} = 3300$~kg/m$^3$ and
$\rho_\text{out} = 2670$~kg/m$^3$ on the inner and outer radii, respectively.
So, the constants $a$ and $b$ from Eq.~\ref{eq:density-linear} must be

@@

\begin{equation}
    b = \rho_\text{out} -
    \frac{\rho_\text{out} - \rho_\text{in}}{R_\text{out} - R_\text{in}} R_\text{out},
\end{equation}

\noindent
where $R_\text{out}$ and $R_\text{in}$ are the outer and inner radii of the shell,
respectively.
Because the outer radius of every spherical shell model will be equal to the mean Earth
radius, $R_\text{out} = 6378.137 \text{km}$.

@@

We compute the gravitational potential ($V$) and the vertical component of the gradient
($g_z$) that each shell model generates on each computation grid.
Other components of the gradient are equal to zero outside of the shell due to the
rotational symmetry and are thus omitted from the analysis.
The computations are repeated for values of $D$ ranging from 0.5 to 5 with a step of
0.5.
Then we calculate the absolute difference between these results and the analytical
solution, taking only the maximum absolute difference on each computation grid.
This differences are shown in Fig.~\ref{fig:D-linear}.
Finally, we set the optimal value of $D$ as the minimum value at which the corresponding
error of the numerical approximation is lower than 0.1\%.

We observe from Fig.~\ref{fig:D-linear} that the relative errors for the
potential and the $g_z$ component fall below the 0.1\% threshold at
$D=1$ and $D=2.5$, respectively, for every shell model and computation grid.
These results also show that there's no meaningful dependency between the accuracy of
the method and the thickness of the shell or the height of the computation grid (the
results for the Satellite grid and the Global grid are very similar).

@@

@@

\begin{equation}
    g_z(r) = \frac{V_\text{sh}(r)}{r},
\end{equation}

>>>>

\noindent
has an analytical solutions for the gravitational potential and its vertical derivative
given by Eqs.~\ref{eq:shell-pot-linear} and \ref{eq:shell-gz}.
The values of the angular and linear coefficients ($a$ and $b$)
can be chosen so that the density assumes the values of $\rho_\text{in} = 3300$~kg/m$^3$
and $\rho_\text{out} = 2670$~kg/m$^3$ on the inner ($R_\text{in}$) and outer
($R_\text{out}$) radii of the shell, respectively,

@@

\begin{equation}
    b = \rho_\text{out} -
    \frac{\rho_\text{out} - \rho_\text{in}}{R_\text{out} - R_\text{in}} R_\text{out}.
\end{equation}

@@

The absolute density difference defined on equation~\ref{eq:density-abs-diff} will
always be zero for the linear density case.
As a result, the inequality~\ref{eq:delta-density} will always be satisfied and no
divisions will ever be performed during the density-based discretization.
Therefore, the two dimensional adaptive discretization algorithm is the
only mechanism that controls the accuracy of the numerical integration for the linear
density case.
For this reason, we will ignore the values of $\delta$ and only determine the minimum
value of $D$ needed in order to guarantee an acceptable accuracy.

We compute the gravitational potential ($V$) and its vertical derivative ($g_z$) for
each shell model in Table~\ref{tab:shell-models} on each computation grid in
Table~\ref{tab:grids}.
The horizontal derivatives of the potential are equal to zero outside of the shell due
to the rotational symmetry and are thus omitted from the analysis.
The computations are repeated for values of the distance-size ratio $D$ ranging from 0.5
to 5 in increments of 0.5.
We then calculate the absolute difference between the numerical results and the
analytical solution for the shell.
Fig.~\ref{fig:D-linear} shows the maximum absolute difference for each shell model and
computation grid as a function of $D$.
The differences are relative to the shell value.
Finally, we set the optimal value of $D$ as the smallest value for which the difference
is below 0.1\%.

We observe from Fig.~\ref{fig:D-linear} that the relative errors for the
potential and $g_z$ fall below the 0.1\% threshold at
$D=1$ and $D=2.5$, respectively.
These values are consistent for all shell models and computation grids.

@@

% Brought this down from the results. It fits better in the discussion. I'll find a
% place for it afterwards.
These results also show that there's no meaningful dependency between the accuracy of
the method and the thickness of the shell or the height of the computation grid (the
results for the Satellite grid and the Global grid are very similar).

@@

\begin{equation}
    g_z(r) = \frac{V_\text{sh}(r)}{r},
\label{eq:shell-gz}
\end{equation}
